---
title: "Project1_gmc"
author: "gina"
date: "12/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISwR)
data("secher")
library(HH)
data("birthweight")
birthweight
library(tidyverse)
```

## Join Function
The name of a column was changed so that it may be joined more easily.
```{R}
sech1 <- secher %>% dplyr::select(birthwt = bwt, bpd, ad, no)
head(sech1)
```

## Tidy Data Example
The data was already pretty tidy.  Here is an example of how the pivot_wider() function could be used to tidy data
```{R}
pivgender <- pivot_wider(birthweight, names_from = "gender", values_from = "gender")
head(pivgender)
```

An observation number was given to the birthweight data set to help visualize the joined data set.
```{R}
btwt <- birthweight %>% mutate(ID = row_number())
head(btwt)
```

## Join Data Sets
All of the variables from the two data sets were kept when the data sets were joined by using the full_join() function.  Any NAs were removed from the data.
```{R}
fulldata <- sech1 %>% full_join(y = btwt, by = "birthwt")
fdo <- fulldata %>% na.omit()
head(fdo)
```

## Filter Function
The data set was filtered so that only observations from the male gender were pulled.  Then the data set was filtered so that only obersavtions where the mother had an age of 22 and and gave birth to a female child were pulled.
```{R}
head(fdo %>% filter(gender == "M"))
head(fdo %>% filter(momsage == "22" & gender =="F"))
```

## Arrange Function
The data frame orgnized into a certain order according to  variable using the arrange() function.
```{R}
head(fdo %>% arrange(desc(birthwt), desc(term)))
head(fdo %>% arrange(desc(term), momsage))
```

## Select Function
The select function allows a few columns to be pulled from a data set at a time.
```{R}
head(fdo %>% dplyr::select(birthwt, momsage, ad))
head(fdo %>% dplyr::select(-gender, - no, -ID))
```

## Group By Function
The group_by() function is used to create subsets of the data.  The summarize() funcion allows summary statistics to be calculated.
```{R}
head(fdo %>% group_by(birthwt) %>% summarize(mean_bwt = mean(birthwt, na.rm = T), sd_bwt = sd(birthwt, na.rm=T)))
```

## Plots

```{r}
plot1 <- fdo %>% group_by(birthwt, momsage) %>%
  summarize(mean_bwt=mean(birthwt, na.rm=T),
              sd_bwt=sd(birthwt, na.rm=T), n=n()) %>%
  ggplot(aes(momsage,mean_bwt))+geom_bar(stat="identity")
plot1 + ggtitle("Birthweight") + ylab("Mean Birthweight (g)") + xlab("Mom's Age (years)")
```

```{r}
ggplot(data = fdo, mapping = aes(x = fdo$birthwt, y = fdo$term))+
  geom_point()+
  ggtitle("Birthweight vs Term")+
  xlab("Birthweight")+ ylab("Term")+
  facet_wrap(~fdo$momsage)
```

```{r}
# This plot shows that there is a general trend between the birthweight and head size of the babies
# There is not an apparent trend with the birthweight or head size and what term the babies were born in, some of the low weight babies were born earlier except for one outlier
ggplot(data = fdo, mapping = aes(x = fdo$birthwt, y = fdo$bpd, color = fdo$term))+
  geom_point(stat = "summary", fun.y="mean")+
  ggtitle("Birthweight vs BPD by Term")+
  xlab("Birthweight")+ ylab("BPD")
```

## Summary Functions
The summary functions allows the mean bithrweight, mean BPD, mean AD, mean age of the mother, and mean term that the baby was born to be calculated.xx
```{R}
fdo %>% summarize(mean_bwt = mean(birthwt),mean_bpd = mean(bpd), mean_ad = mean(ad),mean_momsage = mean(momsage), mean_term = mean(term))
```
The summarize function is used here to calculate the standard deviation of each of the previous variables.
```{R}
fdo %>% summarize(sd_bwt = sd(birthwt))
fdo %>% summarize(sd_bwt = sd(birthwt), sd_bpd = sd(bpd), sd_ad = sd(ad), sd_momsage = sd(momsage), sd_term=sd(term))
```
The summarize function is used here to calculate the variation of each of the variables.
```{R}
fdo %>% summarize(var_bwt = var(birthwt),var_bpd = var(bpd), var_ad = var(ad), var_momsage = var(momsage), var_term = var(term))
```
The summarize function is used here to calculate the minimum value of each variable.
```{R}
fdo %>% summarize(min_bwt = min(birthwt), min_bpd = min(bpd), min_ad = min(ad), min_momsage = min(momsage), min_term = min(term))
```
The summarize function is used here to calculate the maximum value of each variable.
```{R}
fdo %>% summarize(max_bwt = max(birthwt), max_bpd = max(bpd), max_ad = max(ad), max_momsage = max(momsage), max_term = max(term))
```
The n_distinct() function count the number of unique values in a variable.
```{R}
fdo %>% summarize(n_distinct(birthwt))
fdo %>% summarize(n_distinct(bpd))
fdo %>% summarize(n_distinct(ad))                 
fdo %>% summarize(n_distinct(momsage))
fdo %>% summarize(n_distinct(term))
```
The cor() function shows the correlation between two variables.
```{R}
fdo %>% summarize(cor(birthwt, term))
fdo %>% summarize(cor(birthwt, bpd))
fdo %>% summarize(cor(birthwt, ad))
fdo %>% summarize(cor(birthwt, momsage))
```

## Covariance
Covariance
```{R}
fdo%>%select_if(is.numeric)%>%cov()
```

```{R}
# corrleation
# the columns with high correlation have values closer to 1, these values represent that the two variables are highly correlated
# for example, babies with high birthweights tend to have larger heads
# there is a negative correlation between birthweight and moms age which means that babies with higher birthweights were tpically born to younger moms
fdo%>%select_if(is.numeric)%>% cor()%>%round(2)
```

```{R}
#eigenvalues
# the eigen vectors show the direction that the data has the most variance
eig1<-fdo%>%select(birthwt, bpd, term, ad, momsage)%>%cor%>%eigen()
eig1
```

```{R}
#PCA
fdoPCA <- fdo %>% select(-no, -ID, -gender)
head(fdoPCA)
```

```{R}
# Scale the data 
fdoPCA_scale<-fdoPCA%>%select_if(is.numeric)%>%scale
PCA <- princomp(fdoPCA_scale)
summary(PCA, loadings=T)
```

```{R}
# convert eigenvector to eigen value
eigvalue <- PCA$sdev^2
```

```{R}
# this shows the proportion of variation of each principle component
variation=round(eigvalue/sum(eigvalue),2) 
```

```{R}
# deciding how many principal components to keep
# Based on the plot, I will keep to PCs
ggplot()+geom_bar(aes(y=variation,x=1:5),stat="identity")+xlab("")+geom_path(aes(y=variation,x=1:5))+
  geom_text(aes(x=1:5,y=variation,label=round(variation,2)),vjust=1:5,col="white",size=5)+
  scale_y_continuous(breaks=seq(0,.6,.2),labels = scales::percent)+
  scale_x_continuous(breaks=1:10)
```

```{R}
# visualizing the data
yeet <-data.frame(PC1=PCA$scores[,1], PC2=PCA$scores[,2])
ggplot(yeet,aes(PC1, PC2))+geom_point()
```